\chapter{Conclusions}
\label{ch:conclusion}

The previous chapters surveyed developmental cell fate decisions through a quantitative lens. Their collective findings offer fresh perspectives into how the structure and function of gene regulatory networks control emergent patterns at the organismal scale. Individually, each chapter advances the fields of systems and developmental biology in unique ways.

Chapter \ref{ch:clones} introduced a computational framework for automated analysis of genetic mosaics; a class of experiments commonly used to probe cell fate decisions \textit{in situ} \cite{Germani2018,Atkins2019}. The framework combines computer vision and statistics to automate the labor-intensive steps of a quantitative work-flow, enabling automated and systematic comparison of cells subject to control and perturbation conditions in an otherwise equivalent background.

Quantitative mosaic analysis itself is not new, nor is it uncommon in high profile publications \cite{Dai2017,Gavish2016,Li2018}. Yet, these studies deploy an irreproducible mix of ad hoc implementations and costly commercial software. Worse still, qualitative analysis pervades less prominent corners of the literature. Contributing a fully automated framework to the open-source ecosystem will make quantitative mosaic analysis accessible to the research community as a whole. A unified framework will also dramatically simplify the reproduction of existing analyses. Chapter \ref{ch:clones} therefore advances the quantification of developmental biology by adding potential for rigor and reproducibility where they are currently lacking.

Chapter \ref{ch:ratio} explored a novel cell fate decision mechanism underlying photoreceptor specification in the \textit{Drosophila} larval eye. Computer vision techniques were used to extract quantitative measurements of Pnt and Yan dynamics from a wealth of confocal microscopy data. Statistical analysis revealed that differentiation is driven by dynamic changes in the ratio between Pnt and Yan, and is agnostic to changes in their absolute concentrations as long as the ratio remains constant. The data therefore provide the first direct evidence that cell fate decisions can be triggered by changes in the relative abundance of separate transcription factors. This finding rebukes the canonical model of photoreceptor specification \cite{Graham2010}. More importantly, it adds a new dimension to our understanding of how multiple transcription factors cooperatively control cell fate decisions, with broad implications for many developmental contexts within and beyond \textit{Drosophila}. 

The ratiometric sensing mechanism identified in Chapter \ref{ch:ratio} adds to a growing body of evidence that cells are able to sense relative changes in the abundance of GRN components \cite{Goentoro2009a,Frick2017}. These discoveries are exciting because relative sensing could potentially isolate cell fate decisions from environmental sources of variation. This is because environmental fluctuations would likely manifest as correlated extrinsic noise that affects all GRN components in a similar manner, causing absolute but not relative concentrations to vary between cells. Relative sensing might then increase fitness in variable environments. 

Regulatory interactions may provide additional layers of stability. Dual-reporter experiments have shown that regulation buffers cell-by-cell differences in yeast gene expression to enhance the precision of decisions to commit to a mating response phenotype \cite{Colman-Lerner2005}. Indeed, Chapter \ref{ch:metabolism} also showed that the microRNA miR-7 buffers Yan expression levels, and by extension R cell fate decisions, against varying biosynthesis capacity. Perhaps future experiments could address how Pnt levels are affected before and after IPC ablation in $yan^{\Delta miR\hyphy 7}$ mutants. 

These observations reflect a central theme of this dissertation; the structure and function of developmental GRNs dictate the robustness of cell fate decisions to environmental variation. Chapter \ref{ch:metabolism} directly embraced this sentiment. It proposed a new theory to explain why the regulatory networks that coordinate cell fate decisions often contain many seemingly redundant repressors acting upon the same target genes. The theory posits that auxiliary negative regulators mitigate erroneous cell fate decisions when cells are rapidly metabolizing, and implies that auxiliary repressors may help GRNs adapt their behavior to environmentally driven variation in metabolism. The central hypothesis is supported by a diverse collection of experiments in which repressor loss-of-function phenotypes are reversed when biosynthesis rates are slowed. A quantitative modeling framework was used to explore the mechanistic origin of this effect, and theoretically demonstrated that auxiliary repressors could avert erroneous decisions by expanding cells capacity to buffer excess protein expression. Quantitative measurements of transcription factor activity confirmed the hypothesized mechanism in vivo.

The theory implies that a novel evolutionary driving force may shape the structure and function of developmental gene regulatory networks. Chapter \ref{ch:introduction} introduced the notion that developmental GRNs guide organisms through a tortuous journey from embryo to adulthood. The journey is not a sprint. Rather, cells must carefully navigate the many twists and turns of developmental programs; rapidly synthesizing GRN components when and where they are needed, then degrading them with equivalent urgency. The protein synthesis and degradation machineries supply the engine and brakes needed to successfully circumvent these obstacles. Individuals stand to benefit from completing the journey quickly because they are generally vulnerable during this time, with little means to defend themselves against predation and other dangers. They could naively swap out the engine for something more potent, but only at the risk of perilous consequences. By simultaneously upgrading the brakes they could realize the best of both worlds. Similarly, organisms could accelerate development by expanding biosynthesis capacity, but the resultant increase in protein expression might misregulate cell fate decisions and drive the emergence of developmental errors. By simultaneously incorporating additional repressors, they could tolerate faster metabolic rates without compromising accuracy. The data presented in Chapter \ref{ch:metabolism} illustrate the inverse perspective; repressors are dispensable when metabolism is slow, much in the way that the original brakes suffice at low speeds. The data therefore imply that auxiliary repressors allow development to proceed more quickly, without any adverse consequences. Beyond reduced vulnerability, shorter generational times also confer a selective advantage by facilitating rapid exploration of the phenotypic landscape, and thus faster adaptation to varied environments. GRNs should therefore be expected to incorporate any topological features that allow development to proceed more quickly. The abundance of redundant regulation found in developmental GRNs certainly appears to support this hypothesis.




Aryal


Csontitute phosphor of dcr1 impacts metabolism and aging in mice


Tollerson2018

Another experimental study of the connection between phenotype penetrance and cell metabolism appeared in PNAS this week. They're looking at how mutations that cause ribosomal queuing are masked under slow growth conditions in E coli. The deleterious effects of these mutations are not truly mitigated under slow-growth (as in our paper), rather the reduced demand for protein synthesis relieves a translational bottleneck. They conclude that elongation factor P enables adaptation to variable growth conditions. They are focused on protein homeostasis, use a less complex organism, and don't attempt to model the underlying mechanism, but many of their figures and overall conclusions are in principle very similar to our study. I presented this during group meeting this morning and Luis agreed it was worth passing along.


Burnaevskiy2018

They do not focus on negative feedback or developmental decisions, but they show (in worms) that organism-scale phenotype penetrance can is driven by variation in metabolic conditions. This is quite similar to how I rationalize the model underlying our metabolism paper.






\section{ Avenues for further exploration }

- Relation to temperature - would require knowledge of temperature/metabolism scaling
- Introduction of simpler model
- Exploration of tradeoff between synthesis and degradation
- Potential for a complete theoretical framework
- Is redundant regulation the only link? What about promoters?






=== EXPAND ===


Chapter \ref{ch:ratio} also used the framework developed in Chapter \ref{ch:clones} to show that regulation stabilizes the Pnt-to-Yan ratio against genetic variation in \textit{pnt} and \textit{yan} copy number. Subsequent R cell fate decisions therefore also appear to be isolated from genetic variation, permitting rapid accumulation of mutations that allow for wider exploration of the fitness landscape. MicroRNAs are known to play a similar role \cite{Cassidy2013,Cassidy2016}.



It is all too easy to coax data into a desired framework
But models should fit the data; not the other way around. 

Core strength of this work is its flexibility
equilibrium model
dynamoc model

yet the newly proposed rationalizing all existing experimental data \cite{Shwartz2013,BoisclairLachance2014,Pelaez2015a}.

A general model based on the statistical physics of transcription factor DNA binding showed that this phenomenon is a natural consequence of competition between the two transcription factors for common binding sites. 

The results also exemplify the importance of both quantitative and dynamic measurements for characterizing developmental systems. 








Each chapter of this dissertation leveraged the power of quantitative measurement and mathematical modeling to tease deeper meaning out of experimental data than would otherwise have been possible using conventional techniques.

ask how cells reliably commit to appropriate fates at the correct time and place.

tease explanations out of experimental data, rebuke existing dogma, propose new mechanisms, and postulate new theories. 

The findings of this dissertation provide a fresh perspective as to how the structure and function of gene regulatory networks control emergent behavior at the organismal scale.

They suggest 

These achievements were attained through principled application of existing knowledge from other technical disciplines.







\section{New computational resources for quantitative biologists}

The research presented in this dissertation spawned several computational tools that may prove valuable to the broader community of quantitative biologists. All of these resources have been made freely available online under open license for unrestricted use and future development. They are published alongside \textbf{FlyEye Silhouette}: \url{http://silhouette.amaral.northwestern.edu}, a GUI-based MacOS application for segmentation, quantification, and annotation of cell nuclei in the \textit{Drosophila} eye imaginal disc that was developed by Helio Tejedor in the Amaral lab at Northwestern University. The list below describes each of the new tools and their high level functions. All are accessible via GitHub repositories mirrored between both \href{https://github.com/sebastianbernasek/}{my personal account} and the \href{https://github.com/amarallab}{Amaral} and \href{https://github.com/bagherilab}{Bagheri} lab accounts. These repositories contain high level API documentation in addition to a series of Jupyter notebooks that walk the user through a series of usage examples. 

\begin{itemize}[leftmargin=*, topsep=10pt, parsep=5pt, partopsep=5pt, itemsep=10pt]
    
  % FLYEYE ANALYSIS
  \item \textbf{FlyEye Analysis}: \url{https://github.com/sebastianbernasek/flyeye}
  \newline
  Python framework for analyzing data generated using FlyEye Silhouette. Core features include inference of cell developmental ages and analysis of the resultant expression dynamics. Also provides tools to quantify expression heterogeneity and spatial patterns.
  
  % FLYEYE ANALYSIS
  \item \textbf{FlyEye Clones}: \url{https://github.com/sebastianbernasek/clones}  
  \newline
  Python framework for automated mosaic analysis of \textit{Drosophila} eye imaginal discs. Core features will be integrated with future versions of FlyEye Silhouette. 
  
  % FLYEYE SYCLONES
  \item \textbf{FlyEye SyClones}: \url{https://github.com/sebastianbernasek/syclones}
  \newline 
  Python framework for generating synthetic microscopy data that mimic key features of mosaic eye imaginal discs.
  
  % PolyTF BINDING
  \item \textbf{PolyTF Binding}: \url{https://github.com/sebastianbernasek/binding}
  \newline 
  Python framework for simulating the equilibrium occupancy of DNA binding sites by one or more polymerizing transcription factors. Utilizes a C backend that efficiently enumerates all possible microstates in a recursive fashion, enabling nested parallelization of the main computational bottleneck. For systems of two or more transcription factors, the implementation confers a major performance advantage over the sequential enumeration strategy proposed by the manuscript that inspired the model \cite{Hope2017}.
  
  % GENESSA
  \item \textbf{GeneSSA}: \url{https://github.com/sebastianbernasek/genessa}
  \newline
  Python framework for exact stochastic simulation of gene regulatory network dynamics \cite{Gillespie1977}. Simulations are executed by a C backend optimized for performance on networks with a narrow scope of pre-defined reaction propensity functions. The limited scope is by design; GeneSSA prioritizes computational efficiency at the expense of flexibility by explicitly hard coding a set of functional forms. This design places GeneSSA among the most performant implementations of the exact stochastic simulation algorithm for several common types of GRNs. The framework may be (and has been) extended to include additional kinetic formulations as they are required.
  
\end{itemize}

\textbf{NOTE}: See Appendix \ref{appendix:resources} for all experimental data and another set of Jupyter notebooks walk the user through reproducing all of the results and figures presented in this dissertation.











\section{ Tools for a quantitative future }

- Need for quantification in each set of experiments
- Danger of 'feedback' diagrams
- Need for quantitative measurements


\section{ Development viewed through the lens of control theory }

- Value of 'dynamics' only perspective
- Discussion on link to molecular mechanisms
- Ratio control circuit





- development of 'penetrance theory' based on analytical framework using moment equations
- rigorous exploration of other control architectures
- quantitative characterization of a well known system (yeast)
-


\section{Robustness as an evolutionary design principle}

	\subsection{Ubiquity of noise}	
	
	\subsection{Scale free ratiometric sensing}
	
	\subsection{Selective pressure on negative regulation}
	
		\paragraph{Evidence of further phenotype suppression (papers sent to Rich)}

\section{Harnessing the power of quantitative models}

	\subsection{Demand for accurate measurements}
	
	\subsection{Knowledge at the expense of molecular detail}
		\paragraph{Dynamic model of ratiometric control.}		
		
	\subsection{Generating hypotheses to guide experiments}
		\paragraph{Cell decisions under variable temperature.}		
		\paragraph{Robustness to loss of promoters.}
		\paragraph{Exploration of simpler systems (yeast mating response).}
		
\section{Outlook for studies of cell decisions}

	\subsection{Open questions for further exploration}
		\paragraph{Robustness of cell decisions to other environmental conditions.}	
					
	\subsection{Consolidating knowledge through collaboration}
	
		\paragraph{Grounding theory with experimental validation.}		
		\paragraph{Open source ecosystem.}


\section{Dynamic perspective of cell behavior}
	\subsection{Dynamics matter}
		\paragraph{Decisions are localized in time.}
	\subsection{Control theoretical insights into cell behavior}
		\paragraph{Yeast papers}
		\paragraph{Khammash papers}


% recent papers on metabolism & phenotypes --- see PNAS presentation and paper to rich/luis
% recent gap genes example, all information encoded in a handful of factors
% emphasis on dynamics - simple model, how does one process scale versus another

\Chapter{Conclusions}

reconcile pnt/yan mechanism with yan over-expression

discuss impact of clones pipeline on pnt/yan study





---------- GET CITES FROM NRSA

Robustness is a fundamental organizational principle underlying the evolution of biological systems7. This need for reliability amidst genetic and environmental perturbations is typically assumed to explain the prevalence of redundancy and feedback within gene regulatory networks8. The proposed research poses a complementary explanation for the ubiquity of these regulatory features; redundant negative feedback confers a selective advantage by enabling rapid growth and development. By exploring this hypothesis, the study promises to identify a novel driving force for increased complexity in gene regulatory network topologies. 



Loss of function mutations affecting regulatory mechanisms have been implicated in the emergence of cancer, autoimmune disease, neurological disorders, diabetes, cardiovascular disease, and morphological abnormalities. Understanding when and how regulation fails is vital to controlling its proper function in humans.

1.	Misregulation of gene expression results in human cancer and disease. Genome-wide association studies have shown non-coding regulatory DNA variation to be strongly associated with human cancer and disease1. Understanding how regulatory variants give rise to disease states is vital to the development and implementation of novel medical strategies. These functional relationships have been explored for a number of transcriptional, post-transcriptional, and epigenetic regulators, but in most cases the precise molecular details remain elusive2–5. This study proposes both mechanistic and phenomenological frameworks for studying the emergence of developmental errors when regulation fails to attenuate transient signals. By studying how developmental processes fail when regulation is perturbed in Drosophila and yeast, the proposed research will advance our understanding of how regulation ensures accurate and timely cellular decisions in all animals.  

In the future, model-informed design of engineered repressors could serve as a platform for tuning signaling dynamics in a manner conducive to proper human development and health.

The prevalence of redundant regulation in gene regulatory networks is often ascribed to a need for robustness against genetic and environmental variability8. These experiments motivate an alternate hypothesis; redundant negative feedback facilitates faster growth and development. While intriguing, the data are limited to fractions of animal populations exhibiting abnormal phenotypes and do not address the underlying mechanism. Any approach toward understanding this phenomenon must bridge the gap between gene expression and phenotypic outcomes with sufficient generality to explain the breadth of surveyed systems. The proposed research plan provides an extension to this hypothesis; redundant negative feedback facilitates faster growth and development by coupling developmental gene expression programs to physiological cell state. My hypothesis is predicated on the notion that cells make mistakes when global rates of transcription and translation outpace regulatory networks’ abilities to attenuate transient signals. From this perspective, redundant repressors provide overflow capacity to buffer against increased protein synthesis, ensuring the timely degradation of transiently expressed proteins when growth and development are fast. Together, increased biosynthesis and sufficient negative feedback would enable development to proceed more quickly without incurring additional errors.

The breadth of experiments in Drosophila points toward a dynamic phenomenon agnostic to the molecular detail of repressors and their targets. To demonstrate the generality of my hypothesis, I propose a coarse-grained framework relating expression of a generic protein to population-wide phenotype distributions that does not depend upon the specifics of each developmental system. My goal is to predict how removal of a negative feedback mechanism causes protein levels within affected cells to differ from their unperturbed counterparts. As nuanced changes in expression yield abnormal morphologies in vivo, differences in protein levels manifest as differences in phenotype penetrance25. I aim to quantify both how different a population of cells behaves when negative feedback is removed and how the magnitude of this difference depends upon biosynthesis rates.

The framework suggests the experiments reflect a general principle of dynamic systems; they are more sensitive to perturbation when internal dynamics are fast. In this case, transcription factor activity is more sensitive to changes in regulation when mRNA and protein biosynthesis rates are high.




Together, these results indicate that tightly coordinated competition between Yan and Pnt lies at the heart of neuronal fate commitment. These observations are consistent with the notion that Yan and Pnt compete for occupancy of shared binding sites in the promoter region of downstream effectors of neuronal differentiation(Gabay et al., 1996; O’Neill et al., 1994). Consequently, altering the expression dynamics of either protein is expected to increase the frequency of erroneous fate commitment, and subsequently increase the likelihood of a roughened eye phenotype. 